---
title: "Maching Learning"
author: "Bj Giudicessi"
date: "May 27, 2016"
output: html_document
---
## Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

### Install Packages to be used. 

```{r setup, warning=FALSE}
library(caret)
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(rattle)
library(randomForest)
library(knitr)
```

## Loading the Data


```{r Loading data}
set.seed(12345)

trainUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

training <- read.csv(url(trainUrl), na.strings=c("NA","#DIV/0!",""))
testing <- read.csv(url(testUrl), na.strings=c("NA","#DIV/0!",""))
```

### Partioning the data
Partioning Training data set into two data sets, 60% for mlTraining, 40% for mlTesting:

```{r partioning training set}
inTrain <- createDataPartition(training$classe, p=0.6, list=FALSE)
mlTraining <- training[inTrain, ]
mlTesting <- training[-inTrain, ]
dim(mlTraining); dim(mlTesting)
```

### Cleaning the data
The following transformations were made to the data:

Transformation: Cleaning NearZeroVariance Variables
```{r Clean}
myDataNZV <- nearZeroVar(mlTraining, saveMetrics=TRUE)
```

Then create another subset without NZV variables:
```{r}
myNZVvars <- names(mlTraining) %in% c("new_window", "kurtosis_roll_belt", "kurtosis_picth_belt",
"kurtosis_yaw_belt", "skewness_roll_belt", "skewness_roll_belt.1", "skewness_yaw_belt",
"max_yaw_belt", "min_yaw_belt", "amplitude_yaw_belt", "avg_roll_arm", "stddev_roll_arm",
"var_roll_arm", "avg_pitch_arm", "stddev_pitch_arm", "var_pitch_arm", "avg_yaw_arm",
"stddev_yaw_arm", "var_yaw_arm", "kurtosis_roll_arm", "kurtosis_picth_arm",
"kurtosis_yaw_arm", "skewness_roll_arm", "skewness_pitch_arm", "skewness_yaw_arm",
"max_roll_arm", "min_roll_arm", "min_pitch_arm", "amplitude_roll_arm", "amplitude_pitch_arm",
"kurtosis_roll_dumbbell", "kurtosis_picth_dumbbell", "kurtosis_yaw_dumbbell", "skewness_roll_dumbbell",
"skewness_pitch_dumbbell", "skewness_yaw_dumbbell", "max_yaw_dumbbell", "min_yaw_dumbbell",
"amplitude_yaw_dumbbell", "kurtosis_roll_forearm", "kurtosis_picth_forearm", "kurtosis_yaw_forearm",
"skewness_roll_forearm", "skewness_pitch_forearm", "skewness_yaw_forearm", "max_roll_forearm",
"max_yaw_forearm", "min_roll_forearm", "min_yaw_forearm", "amplitude_roll_forearm",
"amplitude_yaw_forearm", "avg_roll_forearm", "stddev_roll_forearm", "var_roll_forearm",
"avg_pitch_forearm", "stddev_pitch_forearm", "var_pitch_forearm", "avg_yaw_forearm",
"stddev_yaw_forearm", "var_yaw_forearm")
mlTraining <- mlTraining[!myNZVvars]
#To check the new N?? of observations
dim(mlTraining)
```
Transformation 2: Remove first column of Dataset
```{r}
mlTraining <- mlTraining[c(-1)]
```
Transformation 3: Variables that have more than a 60% NA's are removed from analysis:
```{r}
trainingV3 <- mlTraining #creating another subset to iterate in loop
for(i in 1:length(mlTraining)) { #for every column in the training dataset
        if( sum( is.na( mlTraining[, i] ) ) /nrow(mlTraining) >= .6 ) { #if n?? NAs > 60% of total observations
        for(j in 1:length(trainingV3)) {
            if( length( grep(names(mlTraining[i]), names(trainingV3)[j]) ) ==1)  { #if the columns are the same:
                trainingV3 <- trainingV3[ , -j] #Remove that column
            }   
        } 
    }
}
#To check the new N?? of observations
dim(trainingV3)
```

```{r}
#Seting back to our set:
mlTraining <- trainingV3
rm(trainingV3)
```
Same transformation for mlTesting data:
```{r}
clean1 <- colnames(mlTraining)
clean2 <- colnames(mlTraining[, -58]) #already with classe column removed
mlTesting <- mlTesting[clean1]
testing <- testing[clean2]

#To check the new N?? of observations
dim(mlTesting)
```

```{r}
#To check the new N?? of observations
dim(testing)
```
Coerce data into same data type. To ensure proper functioning of Decision Trees and RandomForest Algorithm with the Test data set provided.
```{r}
for (i in 1:length(testing) ) {
        for(j in 1:length(mlTraining)) {
        if( length( grep(names(mlTraining[i]), names(testing)[j]) ) ==1)  {
            class(testing[j]) <- class(mlTraining[i])
        }      
    }      
}
#And to make sure Coertion really worked, simple smart ass technique:
testing <- rbind(mlTraining[2, -58] , testing) #note row 2 does not mean anything, this will be removed right.. now:
testing <- testing[-1,]
```

## Using Machine Learning algorithms for prediction: Decision Tree
```{r}
modFitA1 <- rpart(classe ~ ., data=mlTraining, method="class")
```
View the decision tree with fancyPlot
```{r}
fancyRpartPlot(modFitA1)
```
Predicting:

```{r}
predictionsA1 <- predict(modFitA1, mlTesting, type = "class")
```
Use confusion Matrix to test results
```{r}
confusionMatrix(predictionsA1, mlTesting$classe)
```
## Using Machine Learning algorithms for prediction: Random Forests
```{r}
modFitB1 <- randomForest(classe ~. , data=mlTraining)
```
Predicting in-sample error
```{r}
predictionsB1 <- predict(modFitB1, mlTesting, type = "class")
```
Using confusion Matrix to test results
```{r}
confusionMatrix(predictionsB1, mlTesting$classe)
```
The Random Forests yielded better Results.

## Generating Files to submit as answers for the Assignment:
Finally, using the provided Test Set out-of-sample error.

For Random Forests we use the following formula, which produced a better prediction in in-sample:
```{r}
predictionsB2 <- predict(modFitB1, testing, type = "class")
```
Generate files with predictions to submit for final test.
```{r}
projectml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

projectml_write_files(predictionsB2)
```
